Sequential Composition for Relaxed Memory: Pomsets with Predicate Transformers

### 2021-07-07 paper/intro.tex
done through section3

In certain cases it is possible to ban concurrent access using separation
\cite{OHearn:2007:RCL:1235896.1236121}, or to accept inefficient
implementation in order to obtain sequential consistency
\cite{DBLP:conf/snapl/MarinoMMNS15}.  When these approaches are not
available, however, we are left with an enormous gap in our understanding of
one of the most basic elements of computing: the humble semicolon.  Until
recently, existing approaches either
\begin{itemize}
\item didn't bother tracking dependencies, allowing ``thin air'' executions
  \cite{Batty:2011:MCC:1926385.1926394},
\item tracked dependencies conservatively, using syntax, requiring inefficient
  implementation of relaxed access
  \cite{DBLP:conf/pldi/LahavVKHD17,Boehm:2014:OGA:2618128.2618134,DBLP:conf/oopsla/VafeiadisN13,DBLP:journals/corr/abs-1804-04214}---while
  this may be fine for C, it's a non-started for safe languages like Java,
  which lack C's ``plain'' access,
\item computed dependencies using non-compositional operational models over
  alternate worlds \cite{Manson:2005:JMM:1047659.1040336,
    DBLP:conf/esop/JagadeesanPR10,
    DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}---these
  models validate many compiler optimizations, but also
  %allow some dubious executions 
  fail to validate temporal safety properties
  (see \textsection\ref{sec:promising}).
\end{itemize}

@@

To manage complexity, we have layered the definitions.  After an overview, we
define sequential dependencies in \textsection\ref{sec:model}.  The next two
sections add concurrency.  In \textsection\ref{sec:mca}, we define \PwTmca{},
which models Java for \mca{} hardware, similar to that of
\citet{DBLP:journals/pacmpl/JagadeesanJR20}.  In \textsection\ref{sec:c11},
we define \PwTc{}, which models \cXI, adapting the approach of
\citet{DBLP:conf/esop/PaviottiCPWOB20}.  In \textsection\ref{sec:additional},
we extend the semantics to include additional features, such as address calculation.
In \textsection{sec:results}, we summarize the results.

\todo{flesh this out}
\begin{itemize}
\item We provide a tool to execute litmus tests for both models.
\item \textsection\ref{sec:drf} We prove \drfsc{} for \PwTmca{}; \MRD{} inherits the result from \rcXI.
\item \textsection\ref{sec:arm} We prove a lowering result for \PwTmca{}.
\item \textsection\ref{sec:additional} We extend the model to include many features.
\end{itemize}

% \subsection{Contributions}

% \todo{}
% We show how predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75} can
% be added to pomsets with
% preconditions~\cite{DBLP:journals/pacmpl/JagadeesanJR20} to create a
% compositional semantics for sequential composition.
\begin{itemize}
\item \S\ref{sec:model} presents the basic model, which satisfies many
  desiderata, but not all.
\item \S\ref{sec:arm} shows two approaches for efficient implementation on
  Arm.  The first uses a suboptimal lowering for acquiring reads.  The second
  uses an optimal lowering, but requires a nontrivial change to the
  definition of sequential composition.
\item \S\ref{sec:additional} generalizes the basic semantics of read and write
  to validate compiler optimizations.
\end{itemize}
% Acknowledgements go here, once we're not double-blinded.
% The definitions in this paper have been formalized in Agda.
Because it is closely related, we expect that the memory-model results of
\cite{DBLP:journals/pacmpl/JagadeesanJR20} apply to our model, including
compositional reasoning for temporal safety properties and {local} \drfsc{}
as in \cite{Dolan:2018:BDR:3192366.3192421,DBLP:conf/ppopp/DongolJR19,promising-ldrf}.
% In \textsection\ref{sec:arm}, we provide an alternative proof strategy for
% efficient compilation to \armeight{}, which improves upon that of
% \cite{DBLP:journals/pacmpl/JagadeesanJR20} by using a recent alternative
% characterization of \armeight{}.

>>>>

In certain cases it is possible to ban concurrent access using separation
\cite{OHearn:2007:RCL:1235896.1236121}, or to accept inefficient
implementation in order to obtain sequential consistency
\cite{DBLP:conf/snapl/MarinoMMNS15}.  When these approaches are not
available, however, we are left with an enormous gap in our understanding of
one of the most basic elements of computing: the humble semicolon.  Until
recently, existing approaches either
\begin{itemize}
\item did not bother tracking dependencies, allowing ``thin air'' executions
  \cite{Batty:2011:MCC:1926385.1926394},
\item tracked dependencies conservatively, using syntax, requiring inefficient
  implementation of relaxed access
  \cite{DBLP:conf/pldi/LahavVKHD17,Boehm:2014:OGA:2618128.2618134,DBLP:conf/oopsla/VafeiadisN13,DBLP:journals/corr/abs-1804-04214}---while
  this may be fine for C, it's a non-started for safe languages like Java,
  which lack C's ``plain'' access,
\item computed dependencies using non-compositional operational models over
  alternate worlds \cite{Manson:2005:JMM:1047659.1040336,
    DBLP:conf/esop/JagadeesanPR10,
    DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}---these
  models validate many compiler optimizations, but also
  %allow some dubious executions 
  fail to validate temporal safety properties
  (see \textsection\ref{sec:promising}).
\end{itemize}

@@

To manage complexity, we have layered the definitions.  After an overview, we
define sequential dependencies in \textsection\ref{sec:model}.  The next two
sections add concurrency.  In \textsection\ref{sec:mca}, we define \PwTmca{},
which provides a Java-like model for \mca{} hardware, similar to that of
\citet{DBLP:journals/pacmpl/JagadeesanJR20}.  In \textsection\ref{sec:c11},
we define \PwTc{}, which models \cXI, adapting the approach of
\citet{DBLP:conf/esop/PaviottiCPWOB20}.  In \textsection\ref{sec:additional},
we extend the semantics to include additional features, such as address
calculation.  In \textsection\ref{sec:results}, we summarize our results.

% \todo{flesh this out}
% \begin{itemize}
% \item We provide a tool to execute litmus tests for both models.
% \item \textsection\ref{sec:sc} We prove \drfsc{} for \PwTmca{}; \MRD{} inherits the result from \rcXI.
% \item \textsection\ref{sec:arm} We prove a lowering result for \PwTmca{}.
% \item \textsection\ref{sec:additional} We extend the model to include many features.
% \end{itemize}

% % \subsection{Contributions}

% % \todo{}
% % We show how predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75} can
% % be added to pomsets with
% % preconditions~\cite{DBLP:journals/pacmpl/JagadeesanJR20} to create a
% % compositional semantics for sequential composition.
% \begin{itemize}
% \item \S\ref{sec:model} presents the basic model, which satisfies many
%   desiderata, but not all.
% \item \S\ref{sec:arm} shows two approaches for efficient implementation on
%   Arm.  The first uses a suboptimal lowering for acquiring reads.  The second
%   uses an optimal lowering, but requires a nontrivial change to the
%   definition of sequential composition.
% \item \S\ref{sec:additional} generalizes the basic semantics of read and write
%   to validate compiler optimizations.
% \end{itemize}
% % Acknowledgements go here, once we're not double-blinded.
% % The definitions in this paper have been formalized in Agda.
% Because it is closely related, we expect that the memory-model results of
% \cite{DBLP:journals/pacmpl/JagadeesanJR20} apply to our model, including
% compositional reasoning for temporal safety properties and {local} \drfsc{}
% as in \cite{Dolan:2018:BDR:3192366.3192421,DBLP:conf/ppopp/DongolJR19,promising-ldrf}.
% % In \textsection\ref{sec:arm}, we provide an alternative proof strategy for
% % efficient compilation to \armeight{}, which improves upon that of
% % \cite{DBLP:journals/pacmpl/JagadeesanJR20} by using a recent alternative
% % characterization of \armeight{}.
